Optimized Multi-Frequency Spectra for Numerical Radiative Transfer

### 2011-08-29 doc/paper/ms.tex
c != inf seems to be working correctly, at least for a constant timestep.

%%%
%% Introduction
%%%
\section{Introduction}
The recent development of advanced numerical methods for radiative transfer has led to a substantial increase in the number of astrophysical problems that are computationally accessible.   Several fully 3D cosmological hydrodynamics codes have implemented adaptive ray-tracing methods (\cite{Abel2002}) for propagating radiation around point sources, including \cite{Wise2011} and others.  Flux-limited diffusion schemes (\cite{Reynolds2009} and others) have begun to surface as well, both in cosmological problems and accretion disk problems.  More recently, hybrid-schemes (VET, Petkova \& Springel, etc.) have begun to make an appearance.  For a review on the state of convergence among these varying numerical methods, see \cite{Iliev2006} and \cite{Iliev2009}.

@@

The consequences of poor frequency resolution have been explored recently by \cite{Wise2011}, who find that for the expansion of an HII region around a $10^5$ K blackbody source in a hydrogen-only medium, the density, temperature, velocity, and ionization profiles are well converged for $n_{\nu} \ge 4$.  Use of a monochromatic spectrum for this problem introduces significant errors since all photons are absorbed at a characteristic column density, whereas multi-frequency treatments achieve some column density dependent behavior, and can thus mimic the behavior of a truly continuous spectrum.  The convergence for this test problem is reassuring, though the prospects for convergence are less clear if one were interested in the absorption processes of multiple chemical species, ionization and heating due to X-rays and their energetic secondary photo-electrons (\cite{Shull1985}; \cite{Furlanetto2010}), or inhomogeneous media.  Such effects are likely important in studies of radiative feedback from stars and active galactic nuclei (AGN), and most certainly in efforts to simulate cosmological reionization.  As a result, an effort must be made to ensure the SEDs used in numerical simulations to represent these sources produce

@@

%%% 
%% Methods & Motivation
%%%
\section{Methods \& Motivation}
The key to our subsequent analysis was inspired by this same convenient feature: if $\Gamma_i$ and $\Heat_i$ can be tabulated for a given source \textit{a-priori}, it means that \textit{precision determination of the ionization state and temperature of the gas surrounding that source relies solely on accurate numerical computation of the photoionization and heating rate integrals in Equations \ref{PhotoionizationRate} and \ref{HeatingRate}}.  So, for a given source, we can compute the values of these integrals numerically for the continuous spectrum, and then as discrete sums given $n_{\nu}$ frequency bins with 
\begin{equation}
    I(E) = \sum_{j=1}^{n_{\nu}} I(E_j) .
\end{equation}

@@

Our radiative transfer scheme is modeled closely after the work of \cite{Fukugita1994} and \cite{Thomas2008}, though for completeness we will reiterate the aspects of these methods most pertinent to the problem at hand.  In general, propagating a radiation field relies on solving differential equations governing the rate of change in the number densities of all ions and the temperature of the gas.  Assuming a medium consisting of hydrogen and helium only, we first solve for the relative abundances of each ion via
\begin{align}
    \frac{d \nHII}{dt} & = \gHI \nHI - \aHII \nel \nHII   \label{HIIRateEquation} \\
    \frac{d \nHeII}{dt} & = \gHeI \nHeI - \bHeI \nel \nHeI + \bHeII \nel \nHeII \label{HeIIRateEquation} \\ \nonumber 
    & - \aHeII \nel \nHeII + \aHeIII \nel \nHeIII - \xHeII \nel \nHeIII  \\
    \frac{d \nHeIII}{dt} & = \gHeII \nHeII - \bHeII \nel \nHeII + \aHeIII \nel \nHeIII . \label{HeIIIRateEquation}
\end{align}
Each of these equations represents the balance of ionizations and recombinations for ion $i = \HII, \HeII, \HeIII$, where $\Gamma_i$ denotes the photoionization rate coefficient, $\alpha_i$ ($\xi_i$) the case-A (dielectric) recombination rate coefficients, and $\beta_i$ the collisional ionization rate coefficients.  In this work, we use the formulae in Appendix B of \cite{Fukugita1994} to compute the values of $\alpha$, $\beta$, and $\xi$.  The photoionization rate coefficients, $\Gamma_i$, are given by
\begin{equation}
    \Gamma_i = \int_{E_i}^{\infty} \sigma_i(E) \spec(E) \frac{dE}{E}    \label{PhotoionizationRate}
\end{equation}
where $E_i$ and $\sigma_i$ are the ionization threshold and bound-free absorption cross section for species $i$, respectively.  $\spec$ is the spectral intensity, which we write as
\begin{equation}
    \spec(E, r) = \frac{\Lbol}{4\pi r^2} I(E) e^{-\tau(E, r)} .
\end{equation}
Here, $r$ is the distance from the source, $I(E)$ is the normalized spectral energy distribution (SED), satisfying
\begin{equation}
    \int_E I(E) dE = 1 ,
\end{equation}
and the optical depth $\tau(E, r)$ is given as a sum over absorbing species,
\begin{align}
    \tau(E, r) & = \sum_i \int_r \sigma_i(E) n_i(r) dr \nonumber \\
               & = \sum_i \sigma_i(E) \ncol(r)
\end{align}
where $\ncol(r)$ is the column density of species $i$ at distance $r$ from the source.
We calculate the bound-free absorption cross-sections, $\sigma_i$, using the fits of \cite{Verner1996}.  At each time-step, we also solve for the temperature evolution, which is given by
\begin{align}
    \frac{3}{2}\frac{d}{dt}\left(\frac{\kB T \nb}{\mu}\right) & = \fheat \sum_i n_i \Heat_i \nonumber \\
    & - \sum_i \zeta_i n_e n_i - \sum_i \eta_i n_e n_i - \sum_i \psi_i n_e n_i \label{TemperatureEvolution} 
\end{align}
where $\Heat_i$ is the photo-electric heating rate, given by
\begin{equation}
    \Heat_i = \int_{E_i}^{\infty} \sigma_i(E) \spec(E, r)(E - E_i) \frac{dE}{E} \label{HeatingRate} ,
\end{equation}    
and $\zeta_i$, $\eta_i$, and $\psi_i$ are the collisional ionization, recombination, and collisional excitation cooling coefficients, respectively.  Additional parameters include $\nb$, the total number density of baryons (check on this), the mean molecular weight $\mu$, Boltzmann's constant $\kB$, the kinetic temperature of the gas $T$, and $\fheat$, the fraction of a photon's original energy deposited as heat (see \cite{Shull1985}, \cite{Furlanetto2010}).  For simplicity, we neglect Hubble cooling due to cosmological expansion, as well as Compton heating/cooling, and cooling by free-free emission.

>>>>

%%%
%% Introduction
%%%
\section{Introduction}
The recent development of advanced numerical methods for radiative transfer has led to a substantial increase in the number of astrophysical problems that are computationally accessible.   Several fully 3D cosmological hydrodynamics codes have implemented adaptive ray-tracing methods (\cite{Abel2002}) for propagating radiation around point sources,  or flux-limited diffusion schemes (\cite{Reynolds2009}) for diffuse radiation fields.  More recently, hybrid-schemes have been implemented in smoothed-particle hydrodynamic codes (e.g. \cite{Petkova2011}).  For a review on the state of convergence among several widely used codes, see \cite{Iliev2006} and \cite{Iliev2009}.

@@

The consequences of poor frequency resolution have been explored recently by \cite{Wise2011}, who find that for the expansion of an HII region around a $10^5$ K blackbody source in a hydrogen-only medium, the density, temperature, velocity, and ionization profiles are well converged for $n_{\nu} \ge 4$.  Use of a monochromatic spectrum for this problem introduces significant errors since all photons are absorbed at a characteristic column density, whereas multi-frequency treatments achieve some column density dependent behavior, and can thus mimic the behavior of a truly continuous spectrum.  The convergence for this test problem is reassuring, though the prospects for convergence are less clear if one were interested in the absorption processes of multiple chemical species, ionization and heating due to X-rays and their energetic secondary photo-electrons (\cite{Shull1985}; \cite{Furlanetto2010}), or inhomogeneous media.  Such effects are likely important in studies of radiative feedback from stars and active galactic nuclei (AGN), and most certainly in efforts to simulate cosmological reionization.  As a result, an effort must be made to ensure that the SEDs used to represent stars in black holes in numerical simulations actually reflect the behavior expected for those sources.  Additional motivation for this work comes from the fact that while spatial, temporal, and mass resolution for a simulation are simple to calculate and choose to optimize for a given problem, frequency resolution is not.  We aim to reduce the guesswork involved in discretizing SEDs, and provide a method for optimization on a problem-by-problem basis, just as one does for the other forms of resolution.

@@

%%% 
%% Multi-Frequency vs. Continuous
%%%
\section{Consequences of Discrete Radiation Fields}
Before we compute the optimal multi-frequency spectra for SEDs commonly used in numerical simulations, we will first demonstrate the difference in results brought about by varying frequency resolution.  To do this, we use \texttt{rt1d}, a 1D radiative transfer code developed for this work that is described in more detail in the appendices.  We choose to use 1D

\section{Methods}
The key to our subsequent analysis was inspired by this same convenient feature: if $\Gamma_i$ and $\Heat_i$ can be tabulated for a given source \textit{a-priori}, it means that \textit{precision determination of the ionization state and temperature of the gas surrounding that source relies solely on accurate numerical computation of the photoionization and heating rate integrals in Equations \ref{PhotoionizationRate} and \ref{HeatingRate}}.  So, for a given source, we can compute the values of these integrals numerically for the continuous spectrum, and then as discrete sums given $n_{\nu}$ frequency bins with 
\begin{equation}
    I(E) = \sum_{j=1}^{n_{\nu}} I(E_j) .
\end{equation}

@@

Our radiative transfer scheme is modeled closely after the work of \cite{Fukugita1994} and \cite{Thomas2008}, though for completeness we will reiterate the aspects of these methods most pertinent to the problem at hand.  In general, propagating a radiation field relies on solving differential equations governing the rate of change in the number densities of all ions and the temperature of the gas.  Assuming a medium consisting of hydrogen and helium only, we first solve for the relative abundances of each ion via
\begin{align}
    \frac{d \nHII}{dt} & = \gHI \nHI - \aHII \nel \nHII   \label{HIIRateEquation} \\
    \frac{d \nHeII}{dt} & = \gHeI \nHeI - \bHeI \nel \nHeI + \bHeII \nel \nHeII \label{HeIIRateEquation} \\ \nonumber 
    & - \aHeII \nel \nHeII + \aHeIII \nel \nHeIII - \xHeII \nel \nHeIII  \\
    \frac{d \nHeIII}{dt} & = \gHeII \nHeII - \bHeII \nel \nHeII + \aHeIII \nel \nHeIII . \label{HeIIIRateEquation}
\end{align}
Each of these equations represents the balance of ionizations and recombinations for ion $i = \HII, \HeII, \HeIII$, where $\Gamma_i$ denotes the ionization rate coefficient, $\alpha_i$ ($\xi_i$) the case-A (dielectric) recombination rate coefficients, and $\beta_i$ the collisional ionization rate coefficients.  In this work, we use the formulae in Appendix B of \cite{Fukugita1994} to compute the values of $\alpha$, $\beta$, and $\xi$.  The ionization rate coefficients are composed of terms taking into account photo-ionization, collisional ionization, and secondary ionizations due to energetic photo-electrons.  The photoionization rate coefficients, $\Gamma_i^P$, are given by
\begin{equation}
    \Gamma_i = \int_{E_i}^{\infty} \sigma_i(E) \spec(E) \frac{dE}{E}    \label{PhotoionizationRate}
\end{equation}
where $E_i$ and $\sigma_i$ are the ionization threshold and bound-free absorption cross section for species $i$, respectively.  $\spec$ is the spectral intensity, which we write as
\begin{equation}
    \spec(E, r) = \frac{\Lbol}{4\pi r^2} I(E) e^{-\tau(E, r)} .
\end{equation}
Here, $r$ is the distance from the source, $I(E)$ is the normalized spectral energy distribution (SED), satisfying
\begin{equation}
    \int_E I(E) dE = 1 ,
\end{equation}
and the optical depth $\tau(E, r)$ is given as a sum over absorbing species,
\begin{align}
    \tau(E, r) & = \sum_i \int_r \sigma_i(E) n_i(r) dr \nonumber \\
               & = \sum_i \sigma_i(E) \ncol(r)
\end{align}
where $\ncol(r)$ is the column density of species $i$ at distance $r$ from the source.
We calculate the bound-free absorption cross-sections, $\sigma_i$, using the fits of \cite{Verner1996}.  At each time-step, we also solve for the temperature evolution, which is given by
\begin{align}
    \frac{3}{2}\frac{d}{dt}\left(\frac{\kB T \nb}{\mu}\right) & = \fheat \sum_i n_i \Heat_i \nonumber \\
    & - \sum_i \zeta_i n_e n_i - \sum_i \eta_i n_e n_i - \sum_i \psi_i n_e n_i \label{TemperatureEvolution} 
\end{align}
where $\Heat_i$ is the photo-electric heating rate, given by
\begin{equation}
    \Heat_i = \int_{E_i}^{\infty} \sigma_i(E) \spec(E, r)(E - E_i) \frac{dE}{E} \label{HeatingRate} ,
\end{equation}    
and $\zeta_i$, $\eta_i$, and $\psi_i$ are the collisional ionization, recombination, and collisional excitation cooling coefficients, respectively.  Additional parameters include $\nb$, the total number density of baryons (check on this), the mean molecular weight $\mu$, Boltzmann's constant $\kB$, the kinetic temperature of the gas $T$, and $\fheat$, the fraction of a photon's original energy deposited as heat (see \cite{Shull1985}, \cite{Furlanetto2010}).  For simplicity, we neglect Hubble cooling due to cosmological expansion, as well as Compton heating/cooling, and cooling by free-free emission.
